How to build an homogeneous language for heterogeneous platforms

### 2013-05-30 papers/gpce2013/gpce2013.tex
taking shape

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
maintain. Unfortunately, JavaScript -- which is currently the most supported the action language on Web clients
-- has no static linking system allowing to pick just the functions actually used by a program, forcing clients to
download the whole library even though only a few functions are actually used.

Using the same programming language on both server-side and client-side can improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language can be used on server-side,
and an increasing number of programming languages or compiler back-ends can generate JavaScript code (\eg
Java/GWT~\cite{Chaganti07_GWT}, SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}},
Dart~\cite{Griffith11_Dart}, Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

@@

For instance, because the boundaries of the code sent to the client are less visible when you share code between
client-side and server-side, transitive dependencies may pull a lot of code on the client, causing a high download
overhead. Moreover, generating efficient code for heterogeneous platforms is hard to achieve in an extensible way:
the translation of common abstractions like collections into their native counterpart (JavaScript arrays on
client-side and standard library's collections on server-side) may be hard coded in the compiler, but that would not
scale to handle all the abstractions a complete application may use (\eg HTML fragment definitions, form validation
rules, or even some business data type that may be represented differently for performance reasons).

On one hand, for engineering reasons, developers want to write Web applications using a single language, abstracting
over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

@@

\texttt{js-scala}~\cite{Kossakowski12_JsDESL} is such a compiled embedded DSL defined in Scala that generates
JavaScript code, making it possible to write the client-side code of Web applications using JavaScript. This paper
enriches js-scala in order to solve the problems described above. We use staging (explain) to generate efficient code
for typical abstractions used for Web programming and to generate efficient specialized code for both client and
server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic;
 \item Static linking?
\end{itemize}

@@

\section{Related Work}

Developers express programs using a language that is translated into a form executable on a target platform. This
section presents different approaches for defining cross-platform languages.

\begin{figure}
  \centering
  \includegraphics[width=8cm]{langs.pdf}
  \caption{Language engineering processes}
  \label{langs}
\end{figure}

\subsection{Fat Languages}

One way consists in hard-coding, in the language's compiler, the transformation to each target platform. Figure
\ref{langs} (left) depicts this process. In order to support a feature related to a specific domain, the whole
compiler pipeline (parser, code generator, \etc) may have to be adapted. This approach gives \emph{fat} languages
because a lot of concepts are defined at the language level. Examples of such languages are
Links~\cite{Cooper07_Links}, Opa and Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because
each concept is defined in the compiler, and modifying a compiler requires a high effort. Furthermore, these
languages also require to support common abstraction and composition mechanisms, as general purpose languages do, in
order to make it easier to write real programs. So they need to be at least general purpose languages. It means that
for every problem you have to rewrite a full-featured programming language before addressing the concepts specific to
the problem domain. This approach does not scale.

\subsection{Domain Specific Languages}

Another approach consists in defining several small languages, each one being specific to a given set of concerns,
and then to combine all the source artifacts written with these language into one executable program. This approach
makes it difficult to make the separate languages inter-operate with each other.

\subsection{Thin Languages}

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language. Figure \ref{langs} (middle) depicts this approach. The general purpose language is used as a host language
that does not need to be modified if a new concept is introduced, because concepts are defined as libraries. However,
this approach gives no opportunity to translate a concept efficiently, according to the host platform
characteristics.

\subsection{Deeply Embedded Languages}

Figure \ref{langs} (right) shows a last approach that tries to get the benefits of the previous two approaches,
without their shortcomings.

@@

\section{High-Level Abstractions Generating Efficient (and Heterogeneous) Code}
\label{contribution}

@@

In a Web application, the user interface is defined by a HTML document which can be manipulated by the JavaScript
code to be updated. One typical operation consists in searching some “interesting” element in the document, in order
to extract its content, replace it or listen to user events fired on this element. The standard API provides several
function to search elements in a Web document according to their characteristics such as their attribute values.
Figure~\ref{selectors-api} (wrong reference number) summarizes the available functions and their differences.

\begin{figure}
\label{selectors-api}
\begin{tabular}{| l | p{3cm} |}
\hline
Function & Description \\
\hline
\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
\hline
\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
\hline
\code{querySelectorAll(s)} & All nodes matching the CSS selector \code{s} \\
\hline
\code{getElementsByTagName(n)} & All elements of type \code{n} \\
\hline
\code{getElementsByClassName(c)} & All elements which \code{class} attribute equals to \code{c} \\
\hline
\end{tabular}
\caption{Standard selectors API}
\end{figure}

The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery library (ref and stats) that provides just one function to search for
elements. Listing~\ref{jquery-vanilla-api} shows two equivalent snippets of code, the first one using the native
APIs (a) and the second one using jQuery (b).

\begin{figure}
\begin{lstlisting}[language=JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\caption{(a)}
\begin{lstlisting}
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\caption{(b)}
\label{jquery-vanilla-api}
\end{figure}

@@

Using LMS we are able to provide a simple API (like jQuery) and to generate code using specialized functions (and
hence benefit from their performances), by analyzing the staged program. Listing~\ref{js-scala-selectors} shows how
to implement listing~\ref{jquery-vanilla-api} using js-scala. This listing generates a JavaScript program similar to
listing~\ref{jquery-vanilla-api} (a).

\begin{figure}
\label{js-scala-selectors}
\begin{lstlisting}
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\caption{Selectors in Js-Scala}
\end{figure}

Writing the program with js-scala instead of using a JavaScript library such as jQuery is interesting because the
abstraction for searching elements exist only in the initial program source code, not in the final JavaScript
program. It gives two advantages: (1) the execution of the final program is more performant because of the use of
specialized APIs, (2) the final program’s size is smaller because it does not need to include jQuery.

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider the following typical JavaScript code finding a particular widget in the page and then a particular
button in the widget:

\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget =
  document.querySelector("div.login");
var loginButton =
  loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", handler);
\end{lstlisting}

The native \code{querySelector} method returns \code{null} if no node matched the given selector in the document. If
we run the above code in a page where the widget is not present, it will throw an error and stop further JavaScript
execution. We can write defensive code to handle \code{null} cases, but it leads to very cumbersome code:

\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget =
  document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton =
    loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.
      addEventListener("click", handler);
  }
}
\end{lstlisting}

We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
Here is a Scala listing that uses our DSL (implementation details are given in section \ref{implementation}):

\begin{lstlisting}
for {
  loginWidget <- document.find("div.login")
  loginButton <- loginWidget.find("submit.button")
} loginButton.on(Click) { e => ... }
\end{lstlisting}

The evaluation of the above listing produces a graph of statements from which JavaScript code equivalent to
listing \ref{null-defensive} is generated.

\begin{figure}
\label{dom-api}
\begin{lstlisting}[language=JavaScript]
var articleUi = function (article) {
  var div = document.createElement('div');
  div.setAttribute('class', 'article');
  var span = document.createElement('span');
  var name =
    document.createTextNode(article.name + ': ');
  span.appendChild(name);
  div.appendChild(span);
  var strong = document.createElement('strong');
  var price = document.createTextNode(article.price);
  strong.appendChild(price);
  div.appendChild(strong);
  return div
};
\end{lstlisting}

\begin{lstlisting}[language=HTML]
<div class=article>
  <span>French wine: </span>
  <strong>10</strong>
</div>
\end{lstlisting}
\caption{DOM}
\end{figure}

@@

\subsection{Ad-Hoc Polymorphism}

Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
\emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
the objects. However, modifying existing object prototypes is considered bad
practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
registering supported data types at the beginning of the program execution, as described in section 2.4.3
of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.

We propose to achieve \emph{ad hoc} polymorphism using
typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
compiler, so there is no runtime overhead.

\begin{figure}
\begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
// Interface
case class Show[A](show: Rep[A => Node])

// Polymorphic function
def listWidget[A](items: Rep[List[A]])
      (implicit A: Show[A]): Rep[Node] =
  el("ul")(
    for (item <- items) yield {
      el("li")(A.show(item))
    }
  )

// Type `User`
type User = Record {
  val name: String
  val age: Int
}
// Implementation of Show for a User
implicit val showUser = Show[User] { user =>
  el("span", "class"->"user")(
    user.name + "(" + user.age + " years)"
  )
}

// Main program
def main(users: Rep[List[User]]) = {
  document.body.append(listWidget(users))
}
\end{lstlisting}
\end{figure}

Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
articles.

@@

See http://github.com/js-scala/js-scala.

@@

\begin{figure}
\label{selector-impl}
\begin{lstlisting}
def find(receiver: Exp[SelectorApi],
         selector: Exp[String]) =
  getConstIdentifier(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\caption{Selectors optimization}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
trait JSGenOptionOps extends JSGenEffect {
  val IR: EfffectExp with OptionOpsExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
    rhs match {
      case OptionIsEmpty(o) =>
        emitValDef(sym, quote(o) + " === null")
      case OptionForeach(o, a, b) =>
        stream.println("if (" + quote(o) + " !== null) {")
        emitValDef(a, quote(o))
        emitBlock(b)
        stream.println("}")
      case _ =>
        super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
\end{figure}

@@

\subsection{Ad-Hoc Polymorphism}

@@

Chooze~\footnote{\href{http://chooze.herokuapp.com}{http://chooze.herokuapp.com}} is an existing complete
application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
and look at the results. It contains some JavaScript code to handle the dynamic behavior of the application:
double-posting prevention, dynamic form update and rich interaction with the document.

@@

Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
Chrome v27 Web browser.

\begin{figure}
\label{benchmark}
\centering
\includegraphics[width=8cm]{chooze.png}
\caption{Benchmarks on a real application}
\end{figure}

\section{Conclusion, Future Work}
\label{discussion}

>>>>

The more interactive the application is, the more logic needs to be duplicated between the server-side
and the client-side (explain why?), and the higher is the amount of client-side code. Developers use libraries and
frameworks to get high-level abstractions on client-side, making their code easier to reason about and to
maintain, but also making their code run less efficiently (abstraction penalty).

Using the same programming language on both server-side and client-side could improve the software engineering
process by enabling code reuse between both sides. Incidentally, the JavaScript language -- which is currently the
most supported the action language on Web clients -- can be used on server-side, and an increasing number of
programming languages or compiler back-ends can generate JavaScript code (\eg Java/GWT~\cite{Chaganti07_GWT},
SharpKit\footnote{\href{http://sharpkit.net}{http://sharpkit.net}}, Dart~\cite{Griffith11_Dart},
Kotlin\footnote{\href{http://kotlin.jetbrains.org/}{http://kotlin.jetbrains.org/}},
ClojureScript~\cite{McGranaghan11_ClojureScript}, Fay\footnote{\href{http://fay-lang.org/}{http://fay-lang.org/}},
Haxe~\cite{Cannasse08_HaXe} or Opa\footnote{\href{http://opalang.org/}{http://opalang.org/}}).

@@

Generating efficient code for heterogeneous platforms is hard to achieve in an extensible way: the translation of
common abstractions like collections into their native counterpart (JavaScript arrays on client-side and standard
library's collections on server-side) may be hard coded in the compiler, but that would not scale to handle all the
abstractions a complete application may use (\eg HTML fragment definitions, form validation rules, or even some
business data type that may be represented differently for performance reasons).

On one hand, for engineering reasons, developers want to write Web applications using a single high-level language,
abstracting over the target platforms differences. But on the other hand, for performance reasons, they want to keep
control on the way their code is compiled to each target platform. How to solve this dilemma?

@@

\texttt{js-scala} is such a compiled embedded DSL defined in Scala that generates JavaScript code, making it possible
to write the client-side code of Web applications using JavaScript~\cite{Kossakowski12_JsDESL}. This paper
enriches js-scala in order to solve the problems described above. We use staging (explain?) to generate efficient
code for typical abstractions used for Web programming and to generate efficient specialized code for both client and
server sides for shared abstractions. More precisely, we demonstrate the following features:

\begin{itemize}
 \item An API for searching in the DOM, that exposes a single entry point but that generates code potentially using
more optimized native APIs;
 \item Usage of monads without extra container object creation;
 \item Ability to define DOM fragments using a common language for server-side and client-side, but that generates
code using standard APIs on both server-side and client-side;
 \item Type-directed ad-hoc polymorphism on client-side without runtime dynamic dispatch logic.
\end{itemize}

@@

\section{Related Work (?)}

This section presents different approaches for defining cross-platform programming languages.

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{langs.pdf}
\end{center}
\caption{Language engineering processes}
\label{langs}
\end{figure}

\paragraph{Fat Languages}

The first approach for defining a cross-platform language consists in hard-coding, in the compiler, the
code generation scheme of each language feature to each target platform. Figure \ref{langs} (a) depicts this
process. In order to support a feature related to a specific domain, the whole compiler pipeline (parser, code
generator, \etc) may have to be adapted. This approach gives \emph{fat} languages because a lot of concepts are
defined at the language level: general programming concepts such as naming, functions, classes, and more specific
concepts such as HTML fragment definition. Examples of such languages are Links~\cite{Cooper07_Links}, Opa and
Dart~\cite{Griffith11_Dart}. These languages are difficult to extend because each concept is defined in the compiler,
and modifying a compiler requires a high effort. Furthermore, these languages also require to support common
programming abstraction and composition mechanisms, as general purpose languages do. So they usually try to re-invent
the features of general purpose languages. We argue that this approach for defining programming languages is
difficult to scale: for every problem you have to rewrite a full-featured programming language before addressing the
concepts specific to the problem domain.

\paragraph{Domain Specific Languages}

Another approach consists in defining several small independent languages, each one focusing on concerns specific to
a given problem domain, and then to combine all the source artifacts written with these language into one executable
program (ref), as shown in figure~\ref{langs} (b). Defining such a language requires a minimal effort compared to the
previous approach. On the other hand, it is difficult to have interoperability between different DSLs (ref).
(add refs to DSL-based approaches to define Web programming environments)

\paragraph{Thin Languages}

Alternatively, one can define concepts relative to a specific domain as libraries on top of a thin general purpose
language (this is also referred to as \emph{embedded} domain specific languages). Figure \ref{langs} (c) depicts this
approach. The general purpose language is used as a host language and does not need to be modified if a new concept
is introduced, because concepts are defined as pure libraries. However, this approach gives no opportunity to
translate a concept efficiently according to the host platform characteristics. Examples of languages following this
approach are Java/GWT, Kotlin, HaXe and SharpKit.

\paragraph{Deeply Embedded Languages}

The last approach, shown in figure \ref{langs} (d), can be seen as a middle-ground between the two previous
approaches: DSLs are embedded in a host language but use a code generation process. (benefits and limitations of
this approach)

@@

\section{High-Level Abstractions For Client-Side Code}
\label{contribution}

@@

In a Web application, the user interface is defined by a HTML document that can be updated by the JavaScript code.
A typical operation consists in searching some “interesting” element in the document, in order to extract its
content, replace it or listen to user events triggered on it. The standard API provides several functions to search
elements in a HTML document according to their name or attribute values. Figure~\ref{selectors-api} summarizes the
available functions and their differences.

\begin{figure}
\begin{center}
\begin{tabular}{| l | p{3cm} |}
\hline
Function & Description \\
\hline
\code{querySelector(s)} & First element matching the CSS selector \code{s} \\
\hline
\code{getElementById(i)} & Element which attribute \code{id} equals to \code{i} \\
\hline
\code{querySelectorAll(s)} & All elements matching the CSS selector \code{s} \\
\hline
\code{getElementsByTagName(n)} & All elements of type \code{n} \\
\hline
\code{getElementsByClassName(c)} & All elements which \code{class} attribute contains \code{c} \\
\hline
\end{tabular}
\end{center}
\caption{Standard selectors API}
\label{selectors-api}
\end{figure}

The \code{querySelector} and \code{querySelectorAll} are the most general functions, the others handle special
cases. For the developer it is not convenient to have to master several functions performing similar tasks. In fact,
most JavaScript developers use the jQuery
library~\cite{Bibeault08_jQuery}\footnote{According to the following document:
\href{http://trends.builtwith.com/javascript}{ http://trends.builtwith.com/javascript}, jQuery is used by more than
40\% of the top million sites} that provides only one function to search for elements.
Listings \ref{vanilla-selectors} and \ref{jquery-selectors} show two equivalent JavaScript programs, the first one
using the native APIs and the second one using jQuery.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=vanilla-selectors,caption=Selectors in plain JavaScript]
function getWords() {
  var form = document.getElementById('add-user');
  var sections =
    form.getElementsByTagName('fieldset');
  var results = [];
  for (var i = 0 ; i < sections.length ; i++) {
    var words = sections[i]
      .getElementsByClassName('word');
    results[i] = words;
  }
  return results
}
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=jquery-selectors,caption=Selectors in jQuery]
function getWords() {
  var form = $('#add-user');
  var sections = $('fieldset', form);
  return sections.map(function () {
    return $('.word', this)
  })
}
\end{lstlisting}
\end{figure}

@@

\begin{figure}
\begin{lstlisting}[label=js-scala-selectors,caption=Selectors in js-scala]
def getWords() = {
  val form = document.find("#add-user")
  val sections = form.findAll("fieldset")
  sections map (_.findAll(".word"))
}
\end{lstlisting}
\end{figure}

Listing~\ref{js-scala-selectors} shows how to implement listing~\ref{jquery-selectors} using js-scala. We provide two
functions for searching elements: \code{find} to find the first element matching a selector and \code{findAll} to
find all the matching elements. During staging these functions analyze the selector that is passed as parameter and,
when possible, produce code using the specialized API, otherwise they produce code using \code{querySelector} and
\code{querySelectorAll}. Thanks to this optimizations, listing~\ref{js-scala-selectors} generates a JavaScript
program identical to listing~\ref{vanilla-selectors}. Writing the program with js-scala instead of using a JavaScript
library such as jQuery is interesting because the abstraction for searching elements exist only in the initial
program source code, not in the final JavaScript program. It gives two advantages: (1) the execution of the final
program is more performant because of the use of specialized APIs, (2) the final program’s size is smaller because it
does not need to include jQuery.

@@

Null references are a known source of problems in programming languages~\cite{Hoare09_Null,Nanda09_Null}. For
example, consider listing \ref{null-unsafe} finding a particular widget in the page and then a particular
button in the widget. The native \code{querySelector} method returns \code{null} if no node matched the given
selector in the document. If we run this code in a page where the widget is not present, it will throw an error
and stop further JavaScript execution. Listing \ref{null-defensive} shows defensive code to handle \code{null}
cases, but it leads to very cumbersome code.

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=null-unsafe,caption=Unsafe code]
var loginWidget =
  document.querySelector("div.login");
var loginButton =
  loginWidget.querySelector("button.submit");
loginButton.addEventListener("click", handler);
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=JavaScript,label=null-defensive,caption=Defensive programming to handle null references]
var loginWidget =
  document.querySelector("div.login");
if (loginWidget !== null) {
  var loginButton =
    loginWidget.querySelector("button.submit");
  if (loginButton !== null) {
    loginButton.
      addEventListener("click", handler);
  }
}
\end{lstlisting}
\end{figure}

We want to define a DSL that has both the safety and performance of listing \ref{null-defensive} but the
expressiveness of listing \ref{null-unsafe}. We can get safety by wrapping potentially null values of type
\code{Rep[A]} in a container of type \code{Rep[Option[A]]} requiring explicit dereferencing, we can get
expressiveness by using the Scala \code{for} notation for dereferencing, and finally we can get performance by
generating code that does not actually wraps values in a container but instead checks if they are \code{null} or not
when dereferenced. The wrapping container exists only at the stage-level and is removed during the code generation.
Listing \ref{null-js-scala} shows how to write the same program using our DSL (implementation details are
given in section \ref{implementation}). The evaluation of this listing produces a graph of statements from which
JavaScript code equivalent to listing \ref{null-defensive} is generated.

\begin{figure}
\begin{lstlisting}[label=null-js-scala,caption=Handling null references in js-scala]
for {
  loginWidget <- document.find("div.login")
  loginButton <- loginWidget.find("submit.button")
} loginButton.on(Click) { e => ... }
\end{lstlisting}
\end{figure}

\section{High-Level Abstractions Shared on Clients and Servers}

\begin{figure}
\begin{lstlisting}[language=JavaScript,caption=JavaScript DOM API,label=dom-api]
var articleUi = function (article) {
  var div = document.createElement('div');
  div.setAttribute('class', 'article');
  var span = document.createElement('span');
  var name =
    document.createTextNode(article.name + ': ');
  span.appendChild(name);
  div.appendChild(span);
  var strong = document.createElement('strong');
  var price = document.createTextNode(article.price);
  strong.appendChild(price);
  div.appendChild(strong);
  return div
};
\end{lstlisting}
\end{figure}

\begin{figure}
\begin{lstlisting}[language=HTML,caption=HTML,label=html]
<div class=article>
  <span>French wine: </span>
  <strong>10</strong>
</div>
\end{lstlisting}
\end{figure}

@@

% \subsection{Ad-Hoc Polymorphism}
% 
% Because of the dynamically typed nature of JavaScript, when calling a function there is no proper way to select a
% specialized implementation according to the function’s parameters types. JavaScript is only able to dispatch
% according to a method receiver prototype, \eg{} if one writes \code{foo.bar()} the JavaScript runtime will look into
% the prototype of the \code{foo} object for a property named \code{bar} and will call it. So, the only way to achieve
% \emph{ad hoc} polymorphism on JavaScript objects consists in defining the polymorphic function on the prototypes of
% the objects. However, modifying existing object prototypes is considered bad
% practice~\cite{Zakas12_MaintainableJs}. Another way could consist in manually coding the dispatch logic, by
% registering supported data types at the beginning of the program execution, as described in section 2.4.3
% of~\cite{Abelson83_SICP}, but this solution is painful for developers and incurs a performance overhead.
% 
% We propose to achieve \emph{ad hoc} polymorphism using
% typeclasses~\cite{Wadler89_AdhocPolymorphism,Odersky06_Typeclasses,Oliveira10_Typeclasses} so that it supports
% retroactive extension without modifying objects prototypes. The dispatch logic is type-directed and performed by the
% compiler, so there is no runtime overhead.
% 
% \begin{figure}
% \begin{lstlisting}[label=polymorphism,caption=Ad hoc polymorphism using typeclasses]
% // Interface
% case class Show[A](show: Rep[A => Node])
% 
% // Polymorphic function
% def listWidget[A](items: Rep[List[A]])
%       (implicit A: Show[A]): Rep[Node] =
%   el("ul")(
%     for (item <- items) yield {
%       el("li")(A.show(item))
%     }
%   )
% 
% // Type `User`
% type User = Record {
%   val name: String
%   val age: Int
% }
% // Implementation of Show for a User
% implicit val showUser = Show[User] { user =>
%   el("span", "class"->"user")(
%     user.name + "(" + user.age + " years)"
%   )
% }
% 
% // Main program
% def main(users: Rep[List[User]]) = {
%   document.body.append(listWidget(users))
% }
% \end{lstlisting}
% \end{figure}
% 
% Listing \ref{polymorphism} demonstrates how to define a polymorphic \code{listWidget} function that returns a DOM
% tree containing the representation of a list of items. The \code{Show[A]} typeclass defines how to produce a DOM tree
% for a value of type \code{A}. It is used by the \code{listWidget} function to get the DOM fragments of the list
% items. The listing shows how to reuse the same \code{listWidget} function to show a list of users and a list of
% articles.

@@

\footnote{The code is available at \href{http://github.com/js-scala}{http://github.com/js-scala}}

@@

\begin{figure}
\begin{lstlisting}[label=selector-impl,caption=Selectors optimization]
def find(receiver: Exp[Selector],
         selector: Exp[String]) =
  getConstIdentifier(selector) match {
    case Some(id) if receiver == document =>
      DocumentGetElementById(Const(id))
    case _ =>
      SelectorFind(receiver, selector)
  }
\end{lstlisting}
\end{figure}

Listing \ref{selector-impl}

@@

\begin{figure}
\begin{lstlisting}[caption=Null reference handling DSL code generator,label=option-codegen]
trait JSGenOptionOps extends JSGenEffect {
  val IR: OptionOpsExp
  import IR._
  override def emitNode(sym: Sym[Any], rhs: Def[Any]) =
    rhs match {
      case OptionIsEmpty(o) =>
        emitValDef(sym, q" $o === null")
      case OptionForeach(o, a, b) =>
        stream.println(q"if ($o !== null) {")
        emitValDef(a, quote(o))
        emitBlock(b)
        stream.println("}")
      case _ =>
        super.emitNode(sym, rhs)
    }
}
\end{lstlisting}
\end{figure}

@@

@@

Chooze~\footnote{\href{http://chooze.herokuapp.com}{http://chooze.herokuapp.com}} is an existing complete
application for making polls. It allows users to create a poll, define the choice alternatives, share the poll, vote
and look at the results. It contains JavaScript code to handle the dynamic behavior of the application:
double-posting prevention, dynamic form update and rich interaction with the document.

@@

Rather than doing a micro-benchmark focusing on just one abstraction, we performed a global benchmark that is more
likely to reflect the real performances of the application: we simulated user actions on a Web page (2000 clicks on
buttons, triggering a dynamic update of the page and involving the use of the Option monad, the selectors API and
the HTML fragment definition API) and measured the time it took complete them. We only measured the execution time of
client-side code execution. The tests were run on a DELL Latitude E6430 laptop with 8 GB of RAM, on the Google
Chrome v27 Web browser. Figure \ref{benchmark} shows the benchmark results.

\begin{figure}
\centering
\includegraphics[width=8cm]{chooze.png}
\caption{Benchmarks on a real application}
\label{benchmark}
\end{figure}

\section{Conclusion}
\label{discussion}

We implemented a high-level language abstracting over client and server heterogeneity but producing efficient code.
